Our Galactocentric Coordinate System Is Probably Wrong, and Not by a Little

### 2018-12-08 paper/ms.tex
Update paper

\newcommand{\mi}{\texttt{m12i}}
\newcommand{\mf}{\texttt{m12f}}
\newcommand{\mm}{\texttt{m12m}}

@@

In a completely axisymmetric potential, the notion of a global Galactic
midplane is well defined. However, since the Milky Way is not completely
axisymmetric, the global Galactic midplane is to some level of precision not
well defined. The local Galactic midplane has been found by determining the
height at which the number counts of stars reaches a maximum. This local
midplane is then used as the global midplane, an assumption which would be
justified in a completely axisymmetric potential. Assuming a global Galactic
midplane exists, we show in three cosmological zoom-in simulations from the
FIRE collaboration that local determinations of the midplane can vary by
nearly a thin disk scale height. Using standard orbit integration techniques,
we show that such an inaccuracy in the midplane can induce artifical time
evolution in the vertical action that makes it virtually unusable for
individual stars in the disk. The situation is less dire for angular momentum

% Local determinations of the Galactic midplane rely on finding the height at
% which the number count of stars reaches a maximum. This is not guaranteed to
% be

% In an axisymmetric, adiabatic potential, the 6D phase space of stars' positions and velocities can be reduced to a 3D space of invariant actions. We show that if one assumes the wrong galactocentric coordinate system this can induce significant artificial time dependence in the computed actions.

% In an axisymmetric potential the 6D phase space of stars' positions and
% velocities can be reduced to a 3D space of invariant actions. We show that if
% one assumes the wrong galactocentric coordinate system this can induce
% artificial non-axisymmetries which manifest as time dependence in the computed
% actions. However, the Milky Way is not axisymmetric in detail and so there are
% no truly axisymmetric axes. We instead define the  ``optimally axisymmetric''
% axes as the coordinate system which minimizes the time dependence of the
% computed actions. Standard methods for determining the galactocentric
% coordinate system typically involve finding the median height of stars in the
% disk. Using cosmological zoom-in simulations from FIRE-2, we show that the
% optimally axisymmetric coordinate system can differ from standard methods of
% calculating the galactocentric coordinate system by nearly a scale height.
% This work calls into question the use of actions as dynamical invariants
% unless these coordinate system issues can be resolved.

@@

\section{Local Midplane in FIRE} \label{sec:local_fire}
\subsection{Description of Simulation} \label{ssec:cosmozoom}
Here we briefly describe the cosmological zoom-ins used in this work.
Cosmological zoom-ins allow one to simulate a selected region at high
resolution embedded in a low-resolution cosmological background
\citep[e.g.][]{1993ApJ...412..455K,2014MNRAS.437.1894O}. The FIRE
collaboration has simulated a number of Milky Way-mass zoom-ins as part of the
{\em Latte} suite of FIRE-2 simulations
\citep{2016ApJ...827L..23W,2018MNRAS.481.4133G}. We use the three zoom-ins
considered in \citet{2018arXiv180610564S} --- \texttt{m12i}, \texttt{m12f},
\texttt{m12m}, since they have associated mock {\em Gaia} DR2 catalogues and
the $z=0$ snapshots are publicly available.

@@

These halos are uncontaminated and contain gas particles of mass $\sim 7000
\text{--} 20,000\,\Msun$ and star particles of mass $\sim 5000 \text{--} 7000\,
\Msun$, with the lower end coming from stellar evolution
\citep{2018arXiv180610564S}. Softening lengths for dark matter and star
particles are fixed at $112\,\pc$ and $11.2\,\pc$,
respectively.\footnote{This is $2.8$ times the often-quoted
Plummer-equivalent.} The gas softening length is adaptive, but at $z=0$ the
median softening length for gas particles around roughly solar positions
(with cylindrical radii within $500\,\pc$ of $8.2\,\kpc$ and
$\abs{z}<1\,\kpc$) is $98\,\pc$.

% We restrict ourselves to the simulation \texttt{m12i} \citep[first introduced
% in][]{2016ApJ...827L..23W}. For this work, this galaxy is the best-case
% scenario. It does not contain a bar (like \texttt{m12m}) or companion galaxy,
% and by eye is the closest to axisymmetric of the three galaxies presented in
% \citet{2018arXiv180610564S} \citep{2018MNRAS.481.4133G}.

It is important to remember that simulations of Milky Way-mass galaxies are
not perfect representations of the true Milky Way, as discussed in
\citet{2018arXiv180610564S}. For instance, the velocity structure of
\texttt{m12i} is closer to M31 than the Milky Way. However, in this work we
are most interested in the global properties of the \texttt{m12i} potential,
and specifically in deviations from axisymmetry. From this perspective,
\texttt{m12i} is actually far more axisymmetric than we might expect of the
Milky Way. While \texttt{m12i} has prominent spiral arms, it lacks a bar and
a companion like the Large Magellanic Cloud. In future work, these additional
complications also need to be addressed.

\subsection{Local Midplane} \label{sec:local_midplane}
We

@@

\subsection{Cartoon} \label{ssec:cartoon}
We begin with a cartoon illustrating what we mean in Figure~\ref{fig:cartoon}.
The solid gray line indicates a hypothetical ``true'' orbit. The y-axis
corresponds to the vertical height of the orbit and the x-axis to either time
or orbital phase. The dashed gray line is an incorrect midplane which an
observer uses to compute orbits.

@@

\subsection{Cluster Simulations} \label{ssec:cluster_sim}
We simulate a small star cluster of $256$ stars in the frozen $\z=0$
potential of $\texttt{m12i}$. We use the code \texttt{ph4} within the
\texttt{AMUSE} framework \citep{2011ascl.soft07007P, 2013CoPhC.184..456P,
2013A&A...557A..84P} for the cluster integration. We use a softening length
of $0.01\,\pc$ which means we do not correctly solve the internal dynamics of
the cluster, but this is not important for our purposes. To couple the
gravitational forces of the frozen snapshot to the cluster stars, we use the
\texttt{BRIDGE} \citep{2007PASJ...59.1095F}, also implemented in
\texttt{AMUSE}. Our simulation runs for $2\,\Gyr$, allowing us to solve for
$\sim 10$ orbits.

@@

We use the initial position and velocity of two star particles in
$\texttt{m12i}$. These two star particles are randomly chosen from all star
particles in the simulation with the following properties (in the fiducial
coordinate system):
\begin{enumerate}
    \item age in the range $(0.25, 0.75)\,\Gyr$,
    \item cylindrical radius in the range $(7.7, 8.7)\,\kpc$,
    \item height (absolute value) less than $0.5\,\kpc$,
    \item $J_r$ and $J_z$ in the range:
        \subitem $(10, 20)\,\actunit$ for the first star,
        \subitem $(40, 60)\,\actunit$ for the second star.
\end{enumerate}
The ids we choose are \texttt{23693026} for the first star and
\texttt{17012804} for the second star. We refrain from calling these ``low''
or ``high'' action stars since the actions are computed in a presumably
incorrect coordinate system.

>>>>

@@

In a completely axisymmetric potential, the notion of a global Galactic
midplane is well defined. However, since the Milky Way is not completely
axisymmetric, the global Galactic midplane is to some level of precision not
well defined. The local Galactic midplane has been found by determining the
height at which the number counts of stars reaches a maximum. This local
midplane is then used as the global midplane, an assumption which would be
justified in a completely axisymmetric potential. Assuming a global Galactic
midplane exists, we show in three cosmological zoom-in simulations from the
FIRE collaboration that local determinations of the midplane can vary by up to
a thin disk scale height. The effect is more pronounced in galaxies with less
quiescent histories (e.g. Sagittarius-like encounters). We explore one
consequence of such an inaccuracy of the midplane in the context of orbital
actions. Using standard orbit integration techniques, we show that midplane
inaccuracy induces artifical time evolution in the vertical action that makes
it virtually unusable for individual stars in the disk. The situation is less
dire for angular momentum and radial action determinations, though we also
show that inaccuracies in the solar radius can have a similar effect on the
latter. As a result, we question the use of actions as dynamical invariants in
the Milky Way disk.

@@

\section{Local Midplane in FIRE} \label{sec:local_fire}
\subsection{Description of Simulation} \label{ssec:cosmozoom}
Here we briefly describe the cosmological zoom-ins used in this work.
Cosmological zoom-ins allow one to simulate a selected region at high
resolution embedded in a low-resolution cosmological background
\citep[e.g.][]{1993ApJ...412..455K,2014MNRAS.437.1894O}. The FIRE
collaboration has simulated a number of Milky Way-mass zoom-ins as part of the
{\em Latte} suite of FIRE-2 simulations
\citep{2016ApJ...827L..23W,2018MNRAS.481.4133G}. We use the three zoom-ins
considered in \citet{2018arXiv180610564S} --- m12i, m12f, m12m, since they
have associated mock {\em Gaia} DR2 catalogues and the $z=0$ snapshots are
publicly available.

@@

These halos are uncontaminated and contain gas particles of mass $\sim 7000
\text{--} 20,000\,\Msun$ and star particles of mass $\sim 5000 \text{--} 7000\,
\Msun$\footnote{In fact, the simulation m12i contained a gas
splitting bug which causes a small number of gas and star particles to have
larger than typical masses. These constitute only $\sim0.2\%$ of the star
particles in the $z=0$ snapshots, and are only a factor of a few more massive
than typical star particles. Since our results are broadly consistent between
the three simulations, we ignore this minor complication.}, with the lower end
coming from stellar evolution \citep{2018arXiv180610564S}. Softening lengths
for dark matter and star particles are fixed at $112\,\pc$ and $11.2\,\pc$,
respectively.\footnote{This is $2.8$ times the often-quoted
Plummer-equivalent.} The gas softening length is adaptive, but at $z=0$ the
median softening length for gas particles around roughly solar positions (with
cylindrical radii within $500\,\pc$ of $8.2\,\kpc$ and $\abs{z}<1\,\kpc$) is
$98\,\pc$.

In this work, we take the coordinate systems used in
\citet{2018arXiv180610564S} as our fiducial coordinate systems for each
galaxy. The center of the galaxy is found through an iterative ``shrinking
spheres'' method, where the center is calculated for a given radius which is
then reduced by $50\%$ until converging on the galaxy's center. The center of
mass velocity is then determined by all star particles within $15\,\kpc$ of
this center. The galaxy is then rotated onto the principal axis frame as
determined by stars younger than $1\,\Gyr$ inside of the fiducial solar radius
$R_{\odot} = 8.2\,\kpc$.

% We restrict ourselves to the simulation m12i \citep[first introduced
% in][]{2016ApJ...827L..23W}. For this work, this galaxy is the best-case
% scenario. It does not contain a bar (like \texttt{m12m}) or companion galaxy,
% and by eye is the closest to axisymmetric of the three galaxies presented in
% \citet{2018arXiv180610564S} \citep{2018MNRAS.481.4133G}.

It is important to remember that simulations of Milky Way-mass galaxies are
not perfect representations of the true Milky Way, as discussed in
\citet{2018arXiv180610564S}. For instance, the velocity structure of m12i is
closer to M31 than the Milky Way. However, in this work we are most interested
in the global properties of the m12i potential, and specifically in deviations
from axisymmetry. From this perspective, m12i is actually far more
axisymmetric than we might expect of the Milky Way. While m12i has prominent
spiral arms, it lacks a bar and a companion like the Large Magellanic Cloud.
In future work, these additional complications also need to be addressed.

\subsection{Local Midplane} \label{sec:local_midplane}
We now turn to defining the local midplane that an observer might measure if
they were situated in each of these galaxies. We place our imaginary observer
at the solar radius of $8.2\,\kpc$ and azimuth $\phi$, and consider stars
within a cylinder perpendicular to the fiducial disk

For a given azimuth $\phi$ at the solar radius of $8.2\,\kpc$, we take the
median height of all star particles within a cylinder of radius $0.5\,\kpc$
and total height $2\,\kpc$. This median height is taken to be what our
imaginary observer would measure as the local Galactic midplane, and is
reported for each galaxy in Figure~\ref{fig:midplane}. We find that only $10$
iterations of this procedure are necessary for convergence. We bootstrap
resample $1000$ times and report the $95^{\text{th}}$ minus $5^{\text{th}}$
percentile as the dashed-line error bars.

To allow for potential small inaccuracies in the fiducial coordinate systems,
we also subtract the best fit $\text{A} \sin{\phi + \text{B}} + C$ from the
midplane as a function of azimuth. For $\text{A}$, the values are $-165$,
$45$, $9\,\pc$ and for $\text{C}$ the values are $-69$, $19$, and $-18\,\pc$
for m12i, m12f, and m12m, respectively. For the assumed solar radius of
$8.2\,\kpc$, we can approximate the angle offset for the $z$-axis from the
values of $\text{A}$ --- we compute $1.15$, $0.31$, and $0.06\,\deg$ for m12i,
m12f, and m12m. These angle offsets are consistent with the values given in
\citet{2018arXiv180610564S} for the difference between the $z$-axis as defined
by the gas and stars.

Figure~\ref{fig:midplane} shows the inferred midplane our imaginary observer
would make as a function of azimuth for each galaxy. The $90\%$ interquartile
range for each galaxy is $185$, $162$, $84\,\pc$ for m12i, m12f, and m12m.

\begin{figure*}
\plotone{fig/midplane_fit.pdf}
\caption{Caption.}
\label{fig:midplane}
\end{figure*}

Our analysis makes the assumption that a global midplane is well defined.
However, our results are also consistent with the notion that the global
midplane is not a well-defined concept. Furthermore, none of the galaxies chosen in this work have

@@

\subsection{Cartoon} \label{ssec:cartoon}
We begin with a cartoon in Figure~\ref{fig:cartoon} to show that an inaccurate determination of 
The solid gray line indicates a hypothetical ``true'' orbit. The y-axis
corresponds to the vertical height of the orbit and the x-axis to either time
or orbital phase. The dashed gray line is an incorrect midplane which an
observer uses to compute orbits.

@@

\subsection{Cluster Simulations} \label{ssec:cluster_sim}
We simulate a small star cluster of $256$ stars in the frozen $\z=0$
potential of m12i. We use the code \texttt{ph4} within the
\texttt{AMUSE} framework \citep{2011ascl.soft07007P, 2013CoPhC.184..456P,
2013A&A...557A..84P} for the cluster integration. We use a softening length
of $0.01\,\pc$ which means we do not correctly solve the internal dynamics of
the cluster, but this is not important for our purposes. To couple the
gravitational forces of the frozen snapshot to the cluster stars, we use the
\texttt{BRIDGE} \citep{2007PASJ...59.1095F}, also implemented in
\texttt{AMUSE}. Our simulation runs for $2\,\Gyr$, allowing us to solve for
$\sim 10$ orbits.

@@

We use the initial position and velocity of two star particles in
m12i. These two star particles are randomly chosen from all star
particles in the simulation with the following properties (in the fiducial
coordinate system):
\begin{enumerate}
    \item age in the range $(0.25, 0.75)\,\Gyr$,
    \item cylindrical radius in the range $(7.7, 8.7)\,\kpc$,
    \item height (absolute value) less than $0.5\,\kpc$,
    \item $J_r$ and $J_z$ in the range:
        \subitem $(10, 20)\,\actunit$ for the first star,
        \subitem $(40, 60)\,\actunit$ for the second star.
\end{enumerate}
The ids we choose are \texttt{23693026} for the first star and
\texttt{17012804} for the second star. We refrain from calling these ``low''
or ``high'' action stars since the actions are computed in a presumably
incorrect coordinate system.
